---
title: "Zooming GGraph Plots"
subtitle: |
  Some code demonstrating how to zoom into portions of a ggraph.
date: 2021-11-11
categories: [graphing, ggraph, visualization, random-code-snippets]
editor_options: 
  chunk_output_type: console
---


## Inspiration

I was recently working with a **largish** graph that I am using the ggraph package, and I needed to zoom into a sub-region of the graph.

## Setup

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.width = 10, fig.height = 8)
example_df = structure(list(from = c(1L, 2L, 3L, 4L, 5L, 6L, 7L, 7L, 7L, 7L, 
                                     7L, 7L, 7L, 7L, 7L, 7L, 8L, 9L, 10L, 11L, 12L, 12L, 12L, 13L, 
                                     13L, 13L, 13L, 13L, 13L, 13L, 14L, 15L, 16L, 16L, 17L, 17L, 18L, 
                                     18L, 19L, 20L, 21L, 21L, 21L, 21L, 21L, 21L, 21L, 21L, 21L, 21L, 
                                     21L, 21L, 22L, 22L, 6L, 23L, 23L, 23L, 23L, 23L, 23L, 23L, 23L, 
                                     23L, 23L, 23L, 23L, 24L, 24L, 24L, 24L, 24L, 24L, 24L, 24L, 24L, 
                                     24L, 24L, 24L, 25L, 25L, 25L, 25L, 25L, 25L, 25L, 25L, 25L, 25L, 
                                     25L, 25L, 26L, 26L, 26L, 26L, 26L, 26L, 26L, 26L, 26L, 26L, 26L, 
                                     26L, 27L, 27L, 27L, 27L, 27L, 27L, 27L, 27L, 27L, 27L, 27L, 27L, 
                                     28L, 28L, 28L, 28L, 28L, 28L, 28L, 28L, 28L, 28L, 28L, 28L, 29L, 
                                     29L, 29L, 29L, 29L, 29L, 29L, 29L, 29L, 29L, 29L, 29L, 30L, 30L, 
                                     30L, 30L, 30L, 30L, 30L, 30L, 30L, 30L, 30L, 30L, 31L, 31L, 31L, 
                                     31L, 31L, 31L, 31L, 31L, 31L, 31L, 31L, 31L, 32L, 32L, 32L, 32L, 
                                     32L, 32L, 32L, 32L, 32L, 32L, 32L, 32L, 33L, 33L, 33L, 33L, 33L, 
                                     33L, 33L, 33L, 33L, 33L, 33L, 33L, 34L, 34L, 34L, 34L, 34L, 34L, 
                                     34L, 34L, 34L, 34L, 34L, 34L, 35L, 35L, 35L, 35L, 35L, 35L, 35L, 
                                     35L, 35L, 35L, 35L, 35L, 36L, 36L, 36L, 36L, 36L, 36L, 36L, 36L, 
                                     36L, 36L, 36L, 36L, 37L, 37L, 37L, 37L, 37L, 37L, 37L, 37L, 37L, 
                                     37L, 37L, 37L, 38L, 38L, 38L, 38L, 38L, 38L, 38L, 38L, 38L, 38L, 
                                     38L, 38L, 39L, 39L, 39L, 39L, 39L, 39L, 39L, 39L, 39L, 39L, 39L, 
                                     39L, 40L, 40L, 40L, 40L, 40L, 40L, 40L, 40L, 40L, 40L, 40L, 40L, 
                                     41L, 41L, 41L, 41L, 41L, 41L, 41L, 41L, 41L, 41L, 41L, 41L, 42L, 
                                     22L, 22L, 22L, 22L, 22L, 22L, 22L, 22L, 22L, 22L, 22L, 22L, 22L, 
                                     22L, 22L, 22L, 22L, 43L, 43L, 44L, 44L, 45L, 45L, 46L, 46L, 47L, 
                                     47L, 48L, 48L, 49L, 49L, 50L, 50L, 51L, 52L, 52L, 53L, 53L, 54L, 
                                     54L, 55L, 55L, 56L, 57L, 57L, 58L, 58L, 59L, 60L, 61L, 61L, 62L, 
                                     16L, 17L, 18L, 63L, 64L, 65L, 66L, 67L, 68L, 69L, 70L, 71L, 72L, 
                                     73L, 74L, 75L, 76L, 20L, 77L, 77L, 78L, 79L, 79L, 80L, 81L, 82L, 
                                     7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 14L, 15L, 19L, 21L, 21L, 
                                     22L, 83L, 84L, 45L, 46L, 47L, 49L, 50L, 51L, 52L, 53L, 54L, 55L, 
                                     56L, 57L, 58L, 59L, 85L, 48L, 48L, 48L, 48L, 48L, 46L, 46L, 46L, 
                                     46L, 86L, 86L, 13L, 13L, 22L, 7L, 55L, 55L, 55L, 55L, 55L, 51L, 
                                     51L, 51L, 58L, 58L, 58L, 58L, 53L, 53L, 53L, 53L, 49L, 49L, 49L, 
                                     49L, 87L, 87L), to = c(7L, 7L, 7L, 61L, 61L, 61L, 8L, 88L, 89L, 
                                                            90L, 91L, 92L, 93L, 94L, 95L, 96L, 9L, 7L, 9L, 9L, 16L, 17L, 
                                                            18L, 14L, 15L, 16L, 17L, 18L, 19L, 10L, 10L, 10L, 10L, 42L, 10L, 
                                                            42L, 10L, 42L, 10L, 12L, 45L, 46L, 47L, 48L, 49L, 50L, 52L, 53L, 
                                                            54L, 55L, 57L, 58L, 6L, 9L, 9L, 45L, 46L, 47L, 48L, 49L, 50L, 
                                                            52L, 53L, 54L, 55L, 57L, 58L, 45L, 46L, 47L, 48L, 49L, 50L, 52L, 
                                                            53L, 54L, 55L, 57L, 58L, 45L, 46L, 47L, 48L, 49L, 50L, 52L, 53L, 
                                                            54L, 55L, 57L, 58L, 45L, 46L, 47L, 48L, 49L, 50L, 52L, 53L, 54L, 
                                                            55L, 57L, 58L, 45L, 46L, 47L, 48L, 49L, 50L, 52L, 53L, 54L, 55L, 
                                                            57L, 58L, 45L, 46L, 47L, 48L, 49L, 50L, 52L, 53L, 54L, 55L, 57L, 
                                                            58L, 45L, 46L, 47L, 48L, 49L, 50L, 52L, 53L, 54L, 55L, 57L, 58L, 
                                                            45L, 46L, 47L, 48L, 49L, 50L, 52L, 53L, 54L, 55L, 57L, 58L, 45L, 
                                                            46L, 47L, 48L, 49L, 50L, 52L, 53L, 54L, 55L, 57L, 58L, 45L, 46L, 
                                                            47L, 48L, 49L, 50L, 52L, 53L, 54L, 55L, 57L, 58L, 45L, 46L, 47L, 
                                                            48L, 49L, 50L, 52L, 53L, 54L, 55L, 57L, 58L, 45L, 46L, 47L, 48L, 
                                                            49L, 50L, 52L, 53L, 54L, 55L, 57L, 58L, 45L, 46L, 47L, 48L, 49L, 
                                                            50L, 52L, 53L, 54L, 55L, 57L, 58L, 45L, 46L, 47L, 48L, 49L, 50L, 
                                                            52L, 53L, 54L, 55L, 57L, 58L, 45L, 46L, 47L, 48L, 49L, 50L, 52L, 
                                                            53L, 54L, 55L, 57L, 58L, 45L, 46L, 47L, 48L, 49L, 50L, 52L, 53L, 
                                                            54L, 55L, 57L, 58L, 45L, 46L, 47L, 48L, 49L, 50L, 52L, 53L, 54L, 
                                                            55L, 57L, 58L, 45L, 46L, 47L, 48L, 49L, 50L, 52L, 53L, 54L, 55L, 
                                                            57L, 58L, 45L, 46L, 47L, 48L, 49L, 50L, 52L, 53L, 54L, 55L, 57L, 
                                                            58L, 9L, 97L, 98L, 99L, 100L, 101L, 4L, 5L, 102L, 10L, 103L, 
                                                            104L, 105L, 60L, 106L, 107L, 108L, 109L, 13L, 22L, 110L, 83L, 
                                                            21L, 44L, 21L, 44L, 21L, 44L, 21L, 44L, 21L, 44L, 21L, 44L, 44L, 
                                                            21L, 44L, 21L, 44L, 21L, 44L, 21L, 44L, 44L, 21L, 44L, 21L, 44L, 
                                                            44L, 111L, 111L, 112L, 60L, 83L, 83L, 83L, 22L, 22L, 22L, 22L, 
                                                            22L, 22L, 22L, 22L, 22L, 22L, 22L, 22L, 22L, 83L, 83L, 110L, 
                                                            83L, 83L, 110L, 83L, 83L, 22L, 22L, 113L, 114L, 115L, 116L, 117L, 
                                                            118L, 119L, 120L, 121L, 122L, 83L, 83L, 83L, 86L, 2L, 123L, 78L, 
                                                            21L, 84L, 84L, 84L, 84L, 84L, 84L, 84L, 84L, 84L, 84L, 84L, 84L, 
                                                            84L, 84L, 10L, 79L, 124L, 86L, 125L, 84L, 79L, 124L, 86L, 125L, 
                                                            110L, 83L, 42L, 11L, 126L, 111L, 48L, 79L, 124L, 86L, 125L, 79L, 
                                                            86L, 125L, 79L, 124L, 86L, 125L, 79L, 124L, 86L, 125L, 79L, 124L, 
                                                            86L, 125L, 46L, 48L)), class = "data.frame", row.names = c(NA, 
                                                                                                                       -432L))

library(ggraph)
library(tidygraph)
library(igraph)
example_graph = as_tbl_graph(example_df)
```

## Initial View

Let's generate a plot of this to start.

```{r plot_full}
set.seed(1234)
ggraph(example_graph, "graphopt") +
  geom_edge_link(arrow = arrow(length = unit(2, 'mm')),
                end_cap = circle(10, "mm")) +
  geom_node_point()
```

You can see the bottom left section of this graph looks like a mess.
It sure would be awesome if we could zoom into that region, and replot it.

## Find Those Nodes

We will use `igraph`s community detection to break this into community sets and select the correct group.

```{r find_nodes}
walk_membership = cluster_walktrap(example_graph)
walk_communities = membership(walk_membership)
split_comms = split(names(walk_communities), walk_communities)
names(split_comms) = NULL

which_comm = split_comms[[which(purrr::map_lgl(split_comms, ~ "58" %in% .x))]]
```

## Initial Layout

One way to make sure we can get the same layout is to create the layout, which is treated as a graph by ggraph.

```{r initial_layout}
# we set the seed so its the same layout as previous
set.seed(1234)
example_layout = create_layout(example_graph, "graphopt")

ggraph(example_layout) +
  geom_edge_link(arrow = arrow(length = unit(2, 'mm')),
                end_cap = circle(10, "mm")) +
  geom_node_point()
```

Awesome, looks the same as the previous.

## Subset Nodes and Get Range

```{r subset_nodes}
sub_layout = example_layout %>%
  dplyr::filter(name %in% which_comm)
sub_xlim = range(sub_layout$x)
sub_ylim = range(sub_layout$y)
```

## Plot Subset

We would think we should be able to just use the sub-layout to plot the subset of nodes.
Lets try that first.

```{r plot_subset}
ggraph(sub_layout) +
  geom_edge_link(arrow = arrow(length = unit(2, 'mm')),
                end_cap = circle(10, "mm")) +
  geom_node_point()
```

As we can see, this isn't quite right.
We are missing some of the edges.

## Plot Using Subset Ranges

```{r plot_ranges}
ggraph(example_layout) +
  geom_edge_link(arrow = arrow(length = unit(2, 'mm')),
                end_cap = circle(10, "mm")) +
  geom_node_point() +
  coord_cartesian(xlim = sub_xlim, ylim = sub_ylim)
```

Finally!
This is what we wanted.
And we can see why it's so overlapped in the full plot!
It's pretty much a hairball in that part of the graph.
