---
title: "Heatmap Colormaps!"
subtitle: |
  Examples of good colormaps to use for heatmaps, both divergent and incremental.
date: 2022-01-31
categories: [colormaps, visualization, heatmap, random-code-snippets]
---

## Normal Incrementing Data

Nine times out of ten, you probably need just one of the `{viridis}` colormaps for values that are increasing in one direction.
For example, correlation values that are all positive, or abundances from RNA-Seq, etc.
If you want to use them with `{ComplexHeatmap}`, you do so like this:

```{r viridismap, eval = FALSE}
# correlation example for 0 - 1
# imagine cor_vals is a set of correlation values
library(ComplexHeatmap)
cor_start = 0
cor_end = 1
n_break = 20
viridis_map = circlize::colorRamp2(seq(cor_start, 
                                       cor_end, 
                                       length.out = n_break),
                                   viridis::viridis(n_break))

Heatmap(cor_vals, 
        col = viridis_map, 
        ...)
```

## Divergent Data

In this case, you have values where a **central** value is a natural point that you want to highlight things **above** and **below** that value.
Examples include correlations that include negative and positive correlations, log-fold-changes above and below zero.
In this case, I prefer the `{scico}` package with the `vik` colormap.

```{r scicomap, eval = FALSE}
cor_start = -1
cor_end = 1
n_break = 20

scico_map = circlize::colorRamp2(seq(cor_start
                                     cor_end,
                                     length.out = n_break),
                                 scico::scico(n_break, 
                                              palette = "vik"))

Heatmap(cor_vals, 
        col = scico_map,
        ...)
```

## Other Colormaps?

Do you have other suggestions for colormaps, both incremental and divergent?
Please let me know in the comments!
