---
title: "Coloring Dendrogram Edges with ggraph"
subtitle: |
  Here is how I got edges colored in a dendrogram with ggraph. Use "node." in front of the node data column you want.
date: 08-03-2021
categories: [random-code-snippets, visualization, graphing, dendrogram]
bibliography: refs.bib
---

I wanted to color the dendrogram edges according to their class in ggraph, and I was getting stuck because of something that isn't explicitly mentioned in the documentation, but is implied.
**You must use "node." to access the data from the Node Data in the call to `aes(...)`**.

Lets set it up.
We will borrow from the "Edges" vignette in the ggraph package [@ggraphEdgesVignette].

```{r setup_graph}
library(ggraph)
library(tidygraph)
library(purrr)
library(rlang)

set_graph_style(plot_margin = margin(1,1,1,1))
hierarchy <- as_tbl_graph(hclust(dist(iris[, 1:4]))) %>% 
  mutate(Class = map_bfs_back_chr(node_is_root(), .f = function(node, path, ...) {
    if (leaf[node]) {
      as.character(iris$Species[as.integer(label[node])])
    } else {
      species <- unique(unlist(path$result))
      if (length(species) == 1) {
        species
      } else {
        NA_character_
      }
    }
  }))

hierarchy
```

And with this, we can create a dendrogram.

```{r plot_dendrogram}
ggraph(hierarchy, "dendrogram", height = height) +
  geom_edge_elbow()
```

Nice!
But what if we want the leaves colored by which "Class" they belong to?

```{r plot_class}
ggraph(hierarchy, "dendrogram", height = height) +
  geom_edge_elbow2(aes(color = node.Class))
```

Note the differences in this function call compared to the previous:

1. Using `geom_edge_elbow2` instead of `geom_edge_elbow`
1. Using `node.Class`, not just `Class`. 

The second point is really important!
When you look at the `hierarchy` object printed above, the `Class` bit is part of the `Node Data`, which gets identified by `ggraph` by the prefix "node.".

If we don't use `node.Class`, here is the error:

```{r show_error, error = TRUE}
ggraph(hierarchy, "dendrogram", height = height) +
  geom_edge_elbow2(aes(color = Class))
```
