---
title: "Reusing ggplot2 Colors"
subtitle: |
  When you want to reuse ggplot2 default colors across plots.
date: 2021-12-21
categories: [random-code-snippets, ggplot2, visualization]
editor_options: 
  chunk_output_type: console
---

If you are using {ggplot2} a lot in your code, and don't want to bother with a custom color scheme (I mean, there are lots of good options), but you are also using non-ggplot2 type plots, and want to reuse the colors.

I frequently encounter this when I'm doing principal component analysis (PCA) **combined** with heatmaps plotted by `{ComplexHeatmap}`.
I've colored sample groups in the PCA using `{ggplot2}` defaults, and now I want those same colors in the `{ComplexHeatmap}` row or column annotations.

The trick to this is extracting the `palette` from the color definition you want.
For example, very commonly (in biology at least) we might have 2 classes of samples, cases and controls.
So we need 2 colors.

When we generated the {ggplot2} plot, we would do something like this:

```{r example_plot, eval = FALSE}
ggplot(data_frame, aes(x = var1, y = var2, color = groups))
```

To generate the matching colors for something else, we can do:

```{r get_colors, eval = FALSE}
g_colors = scale_color_discrete()$palette(2)
```
